Statistical performance analysis of data-driven neural mass models

### 2014-12-18 manuscript/manuscript.tex
edit to manuscript
edits to paper

\begin{document}

@@

Mesoscopic mathematical models of the brain are becoming increasingly popular as tools for generating hypotheses on the physical principals that govern neurodynamics. Forward models have led to theory on the emergence of the brains rhythms, the mechanisms of anesthetic agents, the brains response to sensory stimuli, short-term memory formation, visual hallucinations and epileptic seizures. Most recently, advances engineering methods and computing resources have expanded the utility of these models as integral elements of frameworks to estimate of system and parametric states that can not be directly measured using electrophysiology. An established framework to reliably estimate physical properties underlying brain dynamics has the potential to revolutionize how we observe, diagnose, treat and cure neurological disorders. For example, in epilepsy the dynamics of hidden system and parametric states are thought to be highly patient-specific and unknown. Thus the ability to estimate these quantities using the framework will enable a greater understanding of individuals pathologies, leading to more tailored therapies and better patient outcomes.  Furthermore, subject-specific models will enable the application of control theory to the neurosciences, which is typically limited to man-made systems.

To date a large body of work has demonstrated validity of these models in-principal, through the ability to reproduce experimental data. Since the parameters of the models are expected to vary across individuals and across disease states, the models must be tuned to individuals to establish clinical utility. A critical steps in establishing robust model fitting techniques is understanding the performance and feasibility of estimation algorithms. This paper introduces engineering tools to neuroscience community for formally accessing feasibility. We also postulate how these tools can be used to increase performance of estimation methods.

Once models have been fit to subjects a final step in assessing validity will be to examine the within subject predictive power. In order for such a framework to be established as a clinical platform further work is required for establishing validity and feasibility for the mesoscopic models.

@@

\paragraph{Neural Mass Model}
\begin{figure}[ht]
  \begin{center}
    \includegraphics{./figures/pdf/SingleMass_plos.pdf}
  \end{center}
  \caption{\emph{\textbf{Neural Mass Model.} The somas of the neural mass are represented by the triangle. The synaptic connections are represented by the circle. The input mean firing rate, $g(v_m(t))$, is convolved with post-synaptic response kernel (PSRK), shown in the inset on the left, to give the mean membrane potential, $v_n(t)$. This is transformed by the sigmoid, shown in the inset on the right, to give the output mean firing rate, $g(v_n(t))$.}}
  \label{fig:SingleNeuralMass}
\end{figure}
Neural mass models map from a mean firing rate of a population of pre-synaptic neurons to a mean membrane potential of a post-synaptic population, which in turn determines that population's output firing rate. A graphical description of a neural mass model can be seen in Figure~\ref{fig:SingleNeuralMass}.

@@

The firing rate of each population is related to the mean membrane potential by a sigmoidal activation function; the logistic function sigmoid is typically used: 
\begin{align}
    g\left(v_n(t)\right) =& \frac{1}{1+\exp{\left(\varsigma_n\left(v_{0n} - v_n(t)\right)\right)}}.
    \label{eq:sigmoid}
\end{align}
The quantity $\varsigma_n$ describes the slope of the sigmoid (approximating the variance of firing thresholds within the populations) and $v_{0n}$ describes the mean firing threshold. An example of the sigmoidal activation function can be seen in the right inset of Figure~\ref{fig:SingleNeuralMass}.

@@

The bound defined in \eqref{eqn:mse_bound} can be used in several ways. First, different models can be compared to test the feasibility of estimating the unknown states from the measurements available. Secondly, different parameter sets can be tested within a single model to assess how the performance of an estimator is expected to change for different types of activity. Thirdly, the efficiency of a given estimation algorithm (e.g. EKF, particle filter) can be assessed against the lower bound. This indicated whether developing progressively advanced estimation algorithms is warranted, a question that has not been considered in the neural estimation literature thus far.

>>>>

\begin{document}
\maketitle

@@

Mesoscopic mathematical models of the brain are becoming increasingly popular as tools for generating hypotheses on the physical principals that govern neurodynamics. Forward models have led to theory on the emergence of the brains rhythms, the mechanisms of anesthetic agents, the brains response to sensory stimuli, short-term memory formation, visual hallucinations and epileptic seizures. Most recently, advances engineering and statistical inference methods and computing resources have expanded the utility of these models as integral elements of frameworks to estimate system states that can not be directly measured using electrophysiology. An established framework to reliably estimate physical properties underlying brain dynamics has the potential to revolutionize how we observe, diagnose, treat and cure neurological disorders. For example, in epilepsy the dynamics of hidden system and parametric states are thought to be highly patient-specific and unknown. Thus the ability to estimate these quantities using the framework will enable a greater understanding of individuals pathologies, leading to more tailored therapies and better patient outcomes.  Furthermore, subject-specific models will enable the application of control theory to the neurosciences, which is typically limited to man-made systems.

To date a large body of work has attempted to demonstrate the validity of these models in-principal, through the ability to qualitatively reproduce experimental data. Since the parameters of the models are expected to vary across individuals and across disease states, the models must be tuned to individuals to establish clinical utility. A critical steps in establishing robust model fitting techniques is understanding the performance and feasibility of estimation algorithms. This paper introduces engineering tools to neuroscience community for formally accessing feasibility. We also postulate how these tools can be used to increase performance of estimation methods that are important for patient-specific modeling. Once models have been fit to subjects, a final step in assessing validity will be to examine the within subject predictive power. In order for such a framework to be established as a clinical platform further work is required for establishing validity and feasibility for the mesoscopic models.

The literature describing mesoscopic neural models is dominated by neural field models and neural mass models. Models of this type

@@

\paragraph{Neural Mass Model}
\begin{figure}[ht]
  \begin{center}
    \includegraphics{./figures/pdf/SingleMass_plos.pdf}
  \end{center}
  \caption{\emph{\textbf{Neural Mass Model.} The somas of the neural mass are represented by the triangle. The synaptic connections are represented by the circle. The input mean firing rate, $g(v_m(t))$, is convolved with post-synaptic response kernel (PSRK), shown in the inset on the left, to give the mean membrane potential, $v_n(t)$. This is transformed by the sigmoid, shown in the inset on the right, to give the output mean firing rate, $g(v_n(t))$.}}
  \label{fig:SingleNeuralMass}
\end{figure}
Neural mass models map from a mean firing rate of a population of pre-synaptic neurons to a mean membrane potential of a post-synaptic population, which in turn determines that population's output firing rate. A graphical description of a single population neural mass model can be seen in Figure~\ref{fig:SingleNeuralMass}, showing the key elements including the post-synaptic response kernel and the sigmoidal activation function.

@@

The firing rate of each population is related to the mean membrane potential by a sigmoidal activation function; the logistic function sigmoid is typically used: 
\begin{align}
    g\left(v_n(t)\right) =& \frac{1}{1+\exp{\left(\varsigma_n\left(v_{0n} - v_n(t)\right)\right)}}.
    \label{eq:sigmoid}
\end{align}
The quantity $\varsigma_n$ describes the slope of the sigmoid (approximating the variance of firing thresholds within the populations) and $v_{0n}$ describes the mean firing threshold relative to a mean rest potential. An example of the sigmoidal activation function can be seen in the right inset of Figure~\ref{fig:SingleNeuralMass}.

@@

The bound defined in \eqref{eqn:mse_bound} can be used in several ways. First, different models can be compared to test the feasibility of estimating the unknown states from the measurements available. Secondly, different parameter sets can be tested within a single model to assess how the performance of an estimator is expected to change for different types of activity. Thirdly, the efficiency of a given estimation algorithm (e.g. EKF, particle filter) can be assessed against the lower bound. The results from this third test will provide an indication on whether developing progressively advanced estimation algorithms is warranted, a question that has not been considered in the neural estimation literature thus far.
